---
title: "DURIN_Gradient_Cleaning_Climate_Sara"
author: "Sonya R. Geange"
format: html
editor: visual
---

## Introduction to Project

Can add a few lines here about what this script is trying to achieve.

## Install and load packages, Import Data

<!--# this is how you add comments -->

```{r load_packages}
#| message: false

# Load the required libraries

# Toinstall packages the first time, run this code
# install.packages("name")

# once installed, run this code
library(tidyverse)
library(readxl)
library(ggplot2)
library(grid)
library(xlsx)
library(RColorBrewer)
library(lme4)
library(lmerTest)
library(emmeans)
library(tidyr)
library(ggthemes)
library(ggpubr)
```

Import data

```{r import_data}

# Import the dataset for the field and lab
Gradient.df <- read_excel("C:/Users/sarag/Uni/DURIN Gradient Project/data/raw_data/Gradient sampling 2024 - DurinV5.xlsx",
                          col_types = c("guess", "text", "text", "text",
                                        "text", "text", "text", "numeric",
                                        "numeric", "numeric", "numeric", "text",
                                        "numeric", "numeric", "numeric", "numeric",
                                        "numeric","numeric", "numeric", "numeric",
                                        "numeric", "numeric", "numeric", "text",
                                        "text", "text", "text", "text", "text", "text", "text", "text"))
Gradient.dfv2 <- read_csv("C:/Users/sarag/Uni/DURIN Gradient Project/data/raw_data/Gradient sampling 2024 - DurinV5.csv", locale=locale(encoding="latin1")
                          )

# Import ANO site and climate data
all_coords <- read.csv("C:/Users/sarag/Uni/DURIN Gradient Project/data/raw_data/all_coords.csv")

# Sampled ANO sites
ANO_sites <- read_excel("C:/Users/sarag/Uni/DURIN Gradient Project/data/raw_data/Gradient sampling 2024 - DurinV5.xlsx",
                        sheet = "ANOSampledSites",
                          col_types = c("guess", "text", "text", "text", "text"))

```

## Data cleaning

```{r four_corners}

gradient.four.corner.df <- Gradient.dfv2 %>%
  # Drop fill the other columns to complete meta-data
  fill(date, .direction = "down") %>%
  fill(ANO_flatID, .direction = "down") %>%
  # Filter to the DURIN 4 Corner sites
  filter(ANO_flatID %in% c("SE_O",
                           "SE_F",
                           "KA_O",
                           "KA_F",
                           "SO_F",
                           "SO_O",
                           "LY_O",
                           "LY_F",
                           "KK_O",
                           "KK_F")) %>%
  # Drop fill only the key columns to help split the dataset
  fill(ANO_pointID, .direction = "down") %>%
  fill(habitat, .direction = "down") %>%
  fill(collector, .direction = "down") %>%
  fill(NiN_type, .direction = "down") %>%
  fill(species, .direction = "down") %>%
  fill(individual_nr, .direction = "down") %>%
  fill(plant_height_cm, .direction = "down") %>%
  fill(length_lb_cm, .direction = "down") %>%
  fill(nr_segments_lb, .direction = "down")  %>%
  # Making sure that the data columns are numeric, Segment column as character
  mutate(across(c(plant_height_cm,
                  length_lb_cm,
                  nr_segments_lb),
                as.numeric)) %>%
  mutate(across(c(
    length_mm,
    nr_leaves,
    nr_scars,
    nr_branch_scars,
    nr_side_branches,
    nr_fruits,
    leaf_weight,
    stem_weight,
    fruit_weight,
    branches_weight,
    sidebranches_weight),
    as.numeric)) %>%
  mutate(across(c(Segment),
                as.character)) %>%
  # Change the spelling of 'KK_F' and 'KK_O' to match correct format
  # During processing KA's and KK's were accidnetly duplicated during data entry of field information
  # Only KK's should be kept as these have field and lab data (weights)
  filter(ANO_flatID %in% c("KK_O", "KK_F",
                           "SE_O", "SE_F",
                           "SO_O", "SO_F",
                           "LY_O", "LY_F")) %>%
  # Rename KK's to KA's to keep naming structure
  mutate(ANO_flatID = recode(ANO_flatID,
                             'KK_F'='KA_F',
                             'KK_O' = 'KA_O')) %>%
  # Change spelling of 'open' to 'Open' to match DURIN formatting standards
  mutate(habitat = recode(habitat,
                          'open'='Open',
                          'forest'='Forest')) %>%
  # Create a new siteID column, and use info in ANO_flatID to allocate sites
  mutate(siteID = recode(ANO_flatID,
                         'SE_O' = 'Senja',
                         'SE_F' = 'Senja',
                         'KA_O' = 'Kautokeino',
                         'KA_F' = 'Kautokeino',
                         'SO_F' = 'Sogndal',
                         'SO_O' = 'Sogndal',
                         'LY_O' = 'Lygra',
                         'LY_F' = 'Lygra')) %>%
  # Change spelling of species and column name to match DURIN data standards
  mutate(speciesID = recode(species,
                            'vv' = 'VV',
                            'vm' = 'VM')) %>%
  # Create column for full species names, and match DURIN data standards
  mutate(species = recode(species,
                          'vv'= 'Vaccinium vitis-idaea',
                          'vm' = 'Vaccinium myrtillus')) %>%
  # Add context for the sites (e.g. Lygra is in the South and by the Coast)
  mutate(geography = factor(siteID, levels = c("Lygra", "Senja", "Sogndal", "Kautokeino"),
                            labels = c("South", "North", "South", "North"))) %>%
  mutate(oceanity = factor(siteID, levels = c("Lygra", "Senja", "Sogndal", "Kautokeino"),
                           labels = c("Coast", "Coast", "Inland", "Inland"))) %>% 
  mutate(
    leaf_stem_ratio_main = leaf_weight / stem_weight,
    total_biomass = rowSums(select(., leaf_weight:sidebranches_weight), na.rm = TRUE))
# add a new column for Leaf/stem ratio and total biomass to be calculated based on existing weight column content
```

```{r climate}

# Create the climate dataset
# Tidy up the dataset names and values
Gradient.df2 <- Gradient.dfv2 %>%
  # Drop fill only the key columns to help split the dataset
  # Date, the ANO_flatID which is a mix of climate, 4corners and VCG coding and in some instance ANO + pointID
  fill(date, .direction = "down") %>%
  fill(ANO_flatID, .direction = "down") %>%
  # filter out (remove) four corner sites
  filter(! ANO_flatID %in% c("SE_O", "SE_F",
                             "KA_O", "KA_F",
                             "KK_O", "KK_F",
                             "LY_O", "LY_F",
                             "SO_O", "SO_F")) %>%
  # Also drop Mika's sites from VCG
  filter(! ANO_pointID %in% c("MK8_4", "MK9_2", "MK8_2",
                              "MK19_1", "MK8_3", "MK8_1",
                              "MK9_4", "MK9_3")) %>%
  # And the sites from Finland (no coords) %>%
  filter(! ANO_flatID %in% c("KIL")) %>%
  # String split ANO_flatID as some instances have the pointID also included
  separate_wider_delim(cols = ANO_flatID,
                       delim = '_',
                       names = c('ANO_flatID', 'ANO_pointID2'),
                       too_few = "align_start") %>%
  # Merge the original ANO_pointID with the newly populated ANO_point2
  mutate(ANO_pointID = coalesce(ANO_pointID, ANO_pointID2)) %>%
  # Drop fill the rest of the columns now the ANO ID's are correct
  fill(ANO_pointID, .direction = "down") %>%
  fill(habitat, .direction = "down") %>%
  fill(collector, .direction = "down") %>%
  fill(NiN_type, .direction = "down") %>%
  fill(species, .direction = "down") %>%
  fill(individual_nr, .direction = "down") %>%
  fill(plant_height_cm, .direction = "down") %>%
  fill(length_lb_cm, .direction = "down") %>%
  fill(nr_segments_lb, .direction = "down")  %>%
  # For external collaborators
  fill(UniqueID, .direction = "down")  %>%
  # Make these columns numeric
  mutate(across(c(
    # Change the plotID codes back to numeric for later on
    ANO_flatID,
    ANO_pointID,
    # Other normal variables
    plant_height_cm,
    length_lb_cm,
    nr_segments_lb,
    length_mm,
    nr_leaves,
    nr_scars,
    nr_branch_scars,
    nr_side_branches,
    nr_fruits,
    leaf_weight,
    stem_weight,
    fruit_weight,
    branches_weight,
    sidebranches_weight),
    as.numeric)) %>%
  # Change spelling of 'open' to 'Open' to match DURIN formatting standards
  mutate(habitat = recode(habitat,
                          'skog' = 'Forest',
                          'open'='Open',
                          'forest'='Forest')) %>%
  # For consistency with 4corners
  # Create a new siteID column, and use info in ANO_flatID to allocate sites
  add_column(siteID = NA) %>%
  # Change spelling of species and column name to match DURIN data standards
  mutate(speciesID = recode(species,
                            'vv' = 'VV',
                            'vm' = 'VM')) %>%
  # Create column for full species names, and match DURIN data standards
  mutate(species = recode(species,
                          'vv'= 'Vaccinium vitis-idaea',
                          'vm' = 'Vaccinium myrtillus')) %>%
  # Remove columns not needed
  select(!ANO_pointID2) %>%
  # Test create new ANO_flatID column removing additional 0 in front
  mutate(ANO_flatID2 = as.numeric(str_remove(ANO_flatID, "^0+")))
```

```{r climate_ANO_IDs}

Gradient.df2 <- Gradient.df2 %>%
  # Next issue is variable ways that the ANO_flatID has been entered (number of digits)
  # With 2, 3 and 4 letter codes we change the Prefix we add
  # With the values above 1000, we don't want the 0's at all
  mutate(ano_punkt = case_when(
    ANO_flatID2 > 0 & ANO_flatID2 <= 99 ~ paste0("ANO00", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    ANO_flatID2 >= 100 & ANO_flatID2 <= 999 ~ paste0("ANO0", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    ANO_flatID2 >= 1000 ~ paste0("ANO", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    TRUE ~ NA))
```

The following block is unfinished, does not run

```{r sites_ANO_IDs}
#| eval: false

# ANO Sites
ANO_sites2 <-  ANO_sites %>%  # Test create new ANO_flatID column removing additional 0 in front
  mutate(ANO_flatID2 = as.numeric(str_remove(ANO_flatID, "^0+")))
# Next issue is variable ways that the ANO_flatID has been entered (number of digits)
# With 2, 3 and 4 letter codes we change the Prefix we add
# With the values above 1000, we don't want the 0's at all
ANO_sites2 <- ANO_sites2 %>% 
  mutate(ano_punkt = case_when(
    ANO_flatID2 > 0 & ANO_flatID2 <= 99 ~ paste0("ANO00", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    ANO_flatID2 >= 100 & ANO_flatID2 <= 999 ~ paste0("ANO0", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    ANO_flatID2 >= 1000 ~ paste0("ANO", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    TRUE ~ NA))
##########################
#TODO:
# Edit Pål's rows, as all plants are listed as individual 1 atm, whereas we need to use the unique ID column to change this.
# Kristine sent through the dataset with the uniqueID, so cross-reference against that

# For LGV and PT, need to check branch numbers vs branch scars.
# For AH, KB, SG, and maybe MK we counted the number of branches removed (base of branch) from the main stem.
# However, LGV and PT counted the number of tips on the branches removed.

# LGV and PT leaf scars vs leaf counts. They have leaf scars in addition to the leaf counts
# AH, KB, SG and maybe MK have removed leaves and counted, then counted all scars remaining on the branch. 
```

```{r climate_ANO}

# Update ano_punkt so only climate gradient sites exist (as they can be matched)
# So all sites with forest/open design (i.e. 4corners and VCG are as NA's, and Finland samples)
Gradient.df2 <- Gradient.df2 %>%
  mutate(ano_punkt = case_when(
    habitat %in% c("Open", "Forest") ~ NA,
    .default = as.character(ano_punkt)
  ))

# Filter the ANO plot co-ordinate and climate data to keep only distinct records for each location
all_coords_distinct <- all_coords %>%
  distinct(ano_punkt, .keep_all = TRUE)

# Create a dataframe just for the sites with coordinates for climate plotting
Gradient.df3 <- Gradient.df2 %>%
  left_join(all_coords_distinct, join_by(ano_punkt)) %>% 
  mutate(
    leaf_stem_ratio_main = leaf_weight / stem_weight,
    total_biomass = rowSums(select(., leaf_weight:sidebranches_weight), na.rm = TRUE))


# Create a dataframe for just the ANO sampled sites, and link to ANO climate data
ANO_site_climate <- ANO_sites %>%
  left_join(all_coords_distinct, join_by(ano_punkt))

```

## Plots

```{r plots_setup}

## Define colour palette (used with scale fill manual)
Habitat <- c("Forest" = "#083508", "Open" = "#589758")
Species <- c("Vaccinium myrtillus" = "#087e8b",   # berry colour option "#323284"
             "Vaccinium vitis-idaea" =  "#eb5461") # berry colour option "#D93137"
siteID <- c("Lygra" = "#4A235A",
            "Sogndal" = "#154360",
            "Senja" = "#C39BD3",
            "Kautokeino" = "#5499C7")
```

Test plots:

- Vegetation types across temperature and precipitation gradient

- Vegetation types across temperature and precipitation gradient, aesthetic changes

- Temperature impact on Plant Height (by vegetation type)

```{r test_plots}

# Do a test plot of climate space we sampled across
(plot <- ggplot(data = Gradient.df3) +
    geom_jitter(aes(x = precipitation_Annual,
                    y = temperature_WQ,
                    colour = hovedtype),
                # Slightly offset the plots so its easier to visualize the samples per site
                height = 0.05, # change is in y-axis values (i.e. temperature)
                width = 25, # change is in x-axis values (i.e. precipitation)
                size = 2))


# Do a test plot of climate space we sampled across
(plot_ANO_climate <- ggplot(data = ANO_site_climate) +
    geom_jitter(aes(x = precipitation_Annual,
                    y = temperature_WQ,
                    colour = hovedtype),
                # Slightly offset the plots so its easier to visualize the samples per site
                height = 0.1, # change is in y-axis values (i.e. temperature)
                width = 25, # change is in x-axis values (i.e. precipitation)
                size = 2) +
    theme_classic() +
    labs(x = "Precipitation (annual, mm/yr)", y = "Temperature (Warmest Quarter)", colour = "Vegetation Type") )

# Test plot of trait against a climate space - temperature
(plot_height_temp <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = plant_height_cm,
                   colour = hovedtype)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Plant Height (cm)", colour = "Vegetation Type") +
    theme_classic2()
)

```

More test plots:

- Variations of trait against a climate space plots

  - leaf weight against temperature (by vegetation type)

  - plant height against temperature (by species)

```{r test_plots_trait/climate}

# repeat test using weight data against temperature
(plot_leafweight_temp <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = leaf_weight,
                   colour = hovedtype)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Leaf Weight (g)", colour = "Vegetation Type") +
    theme_classic2()
)

# check to see if all samples that are weighed so far were included
sum(is.na(Gradient.df3$leaf_weight)) # 996 NAs, explains most of the removed rows
sum(is.na(Gradient.df3$hovedtype)) # 904
sum(is.na(Gradient.df3$hovedtype) & is.na(Gradient.df3$leaf_weight)) # 875 overlap
# only data either missing hovedtype and/or not yet weighed was omitted, all good


# now with species instead of vegetation type
(plot_height_temp2 <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = plant_height_cm,
                   colour = species)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Plant Height (cm)", colour = "Species") +
    theme_classic2()
)


```

Additional test plots

- showing temperature and precipitation across space

```{r test_plots_Sara}

# temperature and precipiation across space
(plot_space_climate <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = latitude,
                   y = longitude,
                   colour = temperature_WQ,
                   size = precipitation_Annual)) +
    scale_colour_gradient(low = "blue", high = "red")+
    labs(x = "Latitude", y = "Longitude", colour = "Temperature") +
    theme_classic2()
)



```

Names of vegetation types

```{r cleaning_vegetation_types}

unique(Gradient.df3$hovedtype)
class(Gradient.df3$hovedtype)
Gradient.df3$hovedtype <- as.factor(Gradient.df3$hovedtype)
levels(Gradient.df3$hovedtype)

Gradient.df3$hovedtype <- factor(Gradient.df3$hovedtype, 
                                 labels = c('Åpen grunnlendt mark', 'Åpen jordvannsmyr', 'Boreal hei', 'Fastmarksskogsmark', 'Fjellhei leside og tundra', 'Grus- og steindominert strand og strandlinje', 'Myr- og sumpskogsmark', 'Nakent berg', 'Nedbørsmyr', 'Rabbe', 'Semi-naturlig eng', 'Skogsmark'))
# renaming 
# having set file encoding option to "latin1" in the beginning when first importing the data helps R interspret the special letters correctly

Gradient.df3$hovedtype <- factor(Gradient.df3$hovedtype,
                                 levels = c('Nakent berg', 'Åpen grunnlendt mark', 'Fjellhei leside og tundra', 'Fastmarksskogsmark', 'Skogsmark', 'Rabbe', 'Grus- og steindominert strand og strandlinje', 'Boreal hei', 'Semi-naturlig eng', 'Åpen jordvannsmyr', 'Myr- og sumpskogsmark', 'Nedbørsmyr'))
# reordering by NiN types, T for Fastmarkssystemer and V for Våtmarkssystemer 

## should Skogsmark also be called Fastmarksskogsmark?
Gradient.df3 <- Gradient.df3 %>% 
  mutate(hovedtype = recode(hovedtype, 'Skogsmark'='Fastmarksskogsmark'))
```
