---
title: "DURIN_Gradient_Cleaning_Climate_Sara"
author: "Sonya R. Geange"
format: html
editor: visual
---

## Introduction to Project

Can add a few lines here about what this script is trying to achieve.

## Install and load packages, Import Data

<!--# this is how you add comments -->

```{r load_packages}
#| message: false

# Load the required libraries

# Toinstall packages the first time, run this code
# install.packages("name")

# once installed, run this code
library(tidyverse)
library(readxl)
library(ggplot2)
library(grid)
library(xlsx)
library(RColorBrewer)
library(lme4)
library(lmerTest)
library(emmeans)
library(tidyr)
library(ggthemes)
library(ggpubr)
```

Import data

```{r import_data}

# Import the dataset for the field and lab
Gradient.df <- read_excel("C:/Users/sarag/Uni/DURIN Gradient Project/data/raw_data/Gradient sampling 2024 - DurinV5.xlsx",
                          col_types = c("guess", "text", "text", "text",
                                        "text", "text", "text", "numeric",
                                        "numeric", "numeric", "numeric", "text",
                                        "numeric", "numeric", "numeric", "numeric",
                                        "numeric","numeric", "numeric", "numeric",
                                        "numeric", "numeric", "numeric", "text",
                                        "text", "text", "text", "text", "text", "text", "text", "text"))
Gradient.dfv2 <- read_csv("C:/Users/sarag/Uni/DURIN Gradient Project/data/raw_data/Gradient sampling 2024 - DurinV5.csv", locale=locale(encoding="latin1")
                          )
# the latin1 encoding introduced a problem here: added names to empty columns, need to be removed
Gradient.dfv2 <- Gradient.dfv2 %>% 
  select(-...33, -...34, -...35, -...36, -...37)

# Import ANO site and climate data
all_coords <- read.csv("C:/Users/sarag/Uni/DURIN Gradient Project/data/raw_data/all_coords.csv")

# Sampled ANO sites
ANO_sites <- read_excel("C:/Users/sarag/Uni/DURIN Gradient Project/data/raw_data/Gradient sampling 2024 - DurinV5.xlsx",
                        sheet = "ANOSampledSites",
                          col_types = c("guess", "text", "text", "text", "text"))

```

## Data cleaning

```{r four_corners}

gradient.four.corner.df <- Gradient.dfv2 %>%
  # Drop fill the other columns to complete meta-data
  fill(date, .direction = "down") %>%
  fill(ANO_flatID, .direction = "down") %>%
  # Filter to the DURIN 4 Corner sites
  filter(ANO_flatID %in% c("SE_O",
                           "SE_F",
                           "KA_O",
                           "KA_F",
                           "SO_F",
                           "SO_O",
                           "LY_O",
                           "LY_F",
                           "KK_O",
                           "KK_F")) %>%
  # Drop fill only the key columns to help split the dataset
  fill(ANO_pointID, .direction = "down") %>%
  fill(habitat, .direction = "down") %>%
  fill(collector, .direction = "down") %>%
  fill(NiN_type, .direction = "down") %>%
  fill(species, .direction = "down") %>%
  fill(individual_nr, .direction = "down") %>%
  fill(plant_height_cm, .direction = "down") %>%
  fill(length_lb_cm, .direction = "down") %>%
  fill(nr_segments_lb, .direction = "down")  %>%
  # Making sure that the data columns are numeric, Segment column as character
  mutate(across(c(plant_height_cm,
                  length_lb_cm,
                  nr_segments_lb),
                as.numeric)) %>%
  mutate(across(c(
    length_mm,
    nr_leaves,
    nr_scars,
    nr_branch_scars,
    nr_side_branches,
    nr_fruits,
    leaf_weight,
    stem_weight,
    fruit_weight,
    branches_weight,
    sidebranches_weight),
    as.numeric)) %>%
  mutate(across(c(Segment),
                as.character)) %>%
  # Change the spelling of 'KK_F' and 'KK_O' to match correct format
  # During processing KA's and KK's were accidnetly duplicated during data entry of field information
  # Only KK's should be kept as these have field and lab data (weights)
  filter(ANO_flatID %in% c("KK_O", "KK_F",
                           "SE_O", "SE_F",
                           "SO_O", "SO_F",
                           "LY_O", "LY_F")) %>%
  # Rename KK's to KA's to keep naming structure
  mutate(ANO_flatID = recode(ANO_flatID,
                             'KK_F'='KA_F',
                             'KK_O' = 'KA_O')) %>%
  # Change spelling of 'open' to 'Open' to match DURIN formatting standards
  mutate(habitat = recode(habitat,
                          'open'='Open',
                          'forest'='Forest')) %>%
  # Create a new siteID column, and use info in ANO_flatID to allocate sites
  mutate(siteID = recode(ANO_flatID,
                         'SE_O' = 'Senja',
                         'SE_F' = 'Senja',
                         'KA_O' = 'Kautokeino',
                         'KA_F' = 'Kautokeino',
                         'SO_F' = 'Sogndal',
                         'SO_O' = 'Sogndal',
                         'LY_O' = 'Lygra',
                         'LY_F' = 'Lygra')) %>%
  # Change spelling of species and column name to match DURIN data standards
  mutate(speciesID = recode(species,
                            'vv' = 'VV',
                            'vm' = 'VM')) %>%
  # Create column for full species names, and match DURIN data standards
  mutate(species = recode(species,
                          'vv'= 'Vaccinium vitis-idaea',
                          'vm' = 'Vaccinium myrtillus')) %>%
  # Add context for the sites (e.g. Lygra is in the South and by the Coast)
  mutate(geography = factor(siteID, levels = c("Lygra", "Senja", "Sogndal", "Kautokeino"),
                            labels = c("South", "North", "South", "North"))) %>%
  mutate(oceanity = factor(siteID, levels = c("Lygra", "Senja", "Sogndal", "Kautokeino"),
                           labels = c("Coast", "Coast", "Inland", "Inland"))) %>% 
  mutate(
    leaf_stem_ratio_main = leaf_weight / stem_weight,
    total_biomass = rowSums(select(., leaf_weight:sidebranches_weight), na.rm = TRUE))
# add a new column for Leaf/stem ratio and total biomass to be calculated based on existing weight column content
```

```{r climate}

# Create the climate dataset
# Tidy up the dataset names and values
Gradient.df2 <- Gradient.dfv2 %>%
  # Drop fill only the key columns to help split the dataset
  # Date, the ANO_flatID which is a mix of climate, 4corners and VCG coding and in some instance ANO + pointID
  fill(date, .direction = "down") %>%
  fill(ANO_flatID, .direction = "down") %>%
  # filter out (remove) four corner sites
  filter(! ANO_flatID %in% c("SE_O", "SE_F",
                             "KA_O", "KA_F",
                             "KK_O", "KK_F",
                             "LY_O", "LY_F",
                             "SO_O", "SO_F")) %>%
  # Also drop Mika's sites from VCG
  filter(! ANO_pointID %in% c("MK8_4", "MK9_2", "MK8_2",
                              "MK19_1", "MK8_3", "MK8_1",
                              "MK9_4", "MK9_3")) %>%
  # And the sites from Finland (no coords) %>%
  filter(! ANO_flatID %in% c("KIL")) %>%
  # String split ANO_flatID as some instances have the pointID also included
  separate_wider_delim(cols = ANO_flatID,
                       delim = '_',
                       names = c('ANO_flatID', 'ANO_pointID2'),
                       too_few = "align_start") %>%
  # Merge the original ANO_pointID with the newly populated ANO_point2
  mutate(ANO_pointID = coalesce(ANO_pointID, ANO_pointID2)) %>%
  # Drop fill the rest of the columns now the ANO ID's are correct
  fill(ANO_pointID, .direction = "down") %>%
  fill(habitat, .direction = "down") %>%
  fill(collector, .direction = "down") %>%
  fill(NiN_type, .direction = "down") %>%
  fill(species, .direction = "down") %>%
  fill(individual_nr, .direction = "down") %>%
  fill(plant_height_cm, .direction = "down") %>%
  fill(length_lb_cm, .direction = "down") %>%
  fill(nr_segments_lb, .direction = "down")  %>%
  # For external collaborators
  fill(UniqueID, .direction = "down")  %>%
  # Make these columns numeric
  mutate(across(c(
    # Change the plotID codes back to numeric for later on
    ANO_flatID,
    ANO_pointID,
    # Other normal variables
    plant_height_cm,
    length_lb_cm,
    nr_segments_lb,
    length_mm,
    nr_leaves,
    nr_scars,
    nr_branch_scars,
    nr_side_branches,
    nr_fruits,
    leaf_weight,
    stem_weight,
    fruit_weight,
    branches_weight,
    sidebranches_weight),
    as.numeric)) %>%
  # Change spelling of 'open' to 'Open' to match DURIN formatting standards
  mutate(habitat = recode(habitat,
                          'skog' = 'Forest',
                          'open'='Open',
                          'forest'='Forest')) %>%
  # For consistency with 4corners
  # Create a new siteID column, and use info in ANO_flatID to allocate sites
  add_column(siteID = NA) %>%
  # Change spelling of species and column name to match DURIN data standards
  mutate(speciesID = recode(species,
                            'vv' = 'VV',
                            'vm' = 'VM')) %>%
  # Create column for full species names, and match DURIN data standards
  mutate(species = recode(species,
                          'vv'= 'Vaccinium vitis-idaea',
                          'vm' = 'Vaccinium myrtillus')) %>%
  # Remove columns not needed
  select(!ANO_pointID2) %>%
  # Test create new ANO_flatID column removing additional 0 in front
  mutate(ANO_flatID2 = as.numeric(str_remove(ANO_flatID, "^0+")))
```

```{r climate_ANO_IDs}

Gradient.df2 <- Gradient.df2 %>%
  # Next issue is variable ways that the ANO_flatID has been entered (number of digits)
  # With 2, 3 and 4 letter codes we change the Prefix we add
  # With the values above 1000, we don't want the 0's at all
  mutate(ano_punkt = case_when(
    ANO_flatID2 > 0 & ANO_flatID2 <= 99 ~ paste0("ANO00", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    ANO_flatID2 >= 100 & ANO_flatID2 <= 999 ~ paste0("ANO0", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    ANO_flatID2 >= 1000 ~ paste0("ANO", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    TRUE ~ NA))
```

The following block is unfinished, does not run

```{r sites_ANO_IDs}
#| eval: false

# ANO Sites
ANO_sites2 <-  ANO_sites %>%  # Test create new ANO_flatID column removing additional 0 in front
  mutate(ANO_flatID2 = as.numeric(str_remove(ANO_flatID, "^0+")))
# Next issue is variable ways that the ANO_flatID has been entered (number of digits)
# With 2, 3 and 4 letter codes we change the Prefix we add
# With the values above 1000, we don't want the 0's at all
ANO_sites2 <- ANO_sites2 %>% 
  mutate(ano_punkt = case_when(
    ANO_flatID2 > 0 & ANO_flatID2 <= 99 ~ paste0("ANO00", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    ANO_flatID2 >= 100 & ANO_flatID2 <= 999 ~ paste0("ANO0", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    ANO_flatID2 >= 1000 ~ paste0("ANO", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    TRUE ~ NA))
##########################
#TODO:
# Edit Pål's rows, as all plants are listed as individual 1 atm, whereas we need to use the unique ID column to change this.
# Kristine sent through the dataset with the uniqueID, so cross-reference against that

# For LGV and PT, need to check branch numbers vs branch scars.
# For AH, KB, SG, and maybe MK we counted the number of branches removed (base of branch) from the main stem.
# However, LGV and PT counted the number of tips on the branches removed.

# LGV and PT leaf scars vs leaf counts. They have leaf scars in addition to the leaf counts
# AH, KB, SG and maybe MK have removed leaves and counted, then counted all scars remaining on the branch. 
```

```{r climate_ANO}

# Update ano_punkt so only climate gradient sites exist (as they can be matched)
# So all sites with forest/open design (i.e. 4corners and VCG are as NA's, and Finland samples)
Gradient.df2 <- Gradient.df2 %>%
  mutate(ano_punkt = case_when(
    habitat %in% c("Open", "Forest") ~ NA,
    .default = as.character(ano_punkt)
  ))

# Filter the ANO plot co-ordinate and climate data to keep only distinct records for each location
all_coords_distinct <- all_coords %>%
  distinct(ano_punkt, .keep_all = TRUE)

# Create a dataframe just for the sites with coordinates for climate plotting
Gradient.df3 <- Gradient.df2 %>%
  left_join(all_coords_distinct, join_by(ano_punkt)) %>% 
  mutate(
    leaf_stem_ratio_main = leaf_weight / stem_weight,
    total_biomass = rowSums(select(., leaf_weight:sidebranches_weight), na.rm = TRUE))


# Create a dataframe for just the ANO sampled sites, and link to ANO climate data
ANO_site_climate <- ANO_sites %>%
  left_join(all_coords_distinct, join_by(ano_punkt))

```

Names of vegetation types

```{r cleaning_vegetation_types}

unique(Gradient.df3$hovedtype)
class(Gradient.df3$hovedtype)
Gradient.df3$hovedtype <- as.factor(Gradient.df3$hovedtype)
levels(Gradient.df3$hovedtype)

Gradient.df3$hovedtype <- factor(Gradient.df3$hovedtype, 
                                 labels = c('Åpen grunnlendt mark', 'Åpen jordvannsmyr', 'Boreal hei', 'Fastmarksskogsmark', 'Fjellhei leside og tundra', 'Grus- og steindominert strand og strandlinje', 'Myr- og sumpskogsmark', 'Nakent berg', 'Nedbørsmyr', 'Rabbe', 'Semi-naturlig eng', 'Skogsmark'))
# renaming 
# having set file encoding option to "latin1" in the beginning when first importing the data helps R interspret the special letters correctly

Gradient.df3$hovedtype <- factor(Gradient.df3$hovedtype,
                                 levels = c('Nakent berg', 'Åpen grunnlendt mark', 'Fjellhei leside og tundra', 'Fastmarksskogsmark', 'Skogsmark', 'Rabbe', 'Grus- og steindominert strand og strandlinje', 'Boreal hei', 'Semi-naturlig eng', 'Åpen jordvannsmyr', 'Myr- og sumpskogsmark', 'Nedbørsmyr'))
# reordering by NiN types, T for Fastmarkssystemer and V for Våtmarkssystemer 

## should Skogsmark also be called Fastmarksskogsmark?
Gradient.df3 <- Gradient.df3 %>% 
  mutate(hovedtype = recode(hovedtype, 'Skogsmark'='Fastmarksskogsmark'))
```

## Plots

```{r plots_setup}

## Define colour palette (used with scale fill manual)
Habitat <- c("Forest" = "#083508", "Open" = "#589758")
Species <- c("Vaccinium myrtillus" = "#087e8b",   # berry colour option "#323284"
             "Vaccinium vitis-idaea" =  "#eb5461") # berry colour option "#D93137"
siteID <- c("Lygra" = "#4A235A",
            "Sogndal" = "#154360",
            "Senja" = "#C39BD3",
            "Kautokeino" = "#5499C7")
```

#### Plots on sites and climate space

1.  Vegetation types across temperature and precipitation gradient

2.  Vegetation types across temperature and precipitation gradient, aesthetic changes

3.  Temperature and precipitation across space (coordinates)

```{r test_plots}

#1 Do a test plot of climate space we sampled across
(plot <- ggplot(data = Gradient.df3) +
    geom_jitter(aes(x = precipitation_Annual,
                    y = temperature_WQ,
                    colour = hovedtype),
                # Slightly offset the plots so its easier to visualize the samples per site
                height = 0.05, # change is in y-axis values (i.e. temperature)
                width = 25, # change is in x-axis values (i.e. precipitation)
                size = 2))


#2 Do a test plot of climate space we sampled across
(plot_ANO_climate <- ggplot(data = ANO_site_climate) +
    geom_jitter(aes(x = precipitation_Annual,
                    y = temperature_WQ,
                    colour = hovedtype),
                # Slightly offset the plots so its easier to visualize the samples per site
                height = 0.1, # change is in y-axis values (i.e. temperature)
                width = 25, # change is in x-axis values (i.e. precipitation)
                size = 2) +
    theme_classic() +
    labs(x = "Precipitation (annual, mm/yr)", y = "Temperature (Warmest Quarter)", colour = "Vegetation Type") )


#3 temperature and precipiation across space
(plot_space_climate <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = longitude,
                   y = latitude,
                   colour = temperature_WQ,
                   size = precipitation_Annual)) +
    scale_colour_gradient(low = "blue", high = "red")+
    labs(x = "Longitude", y = "Latitude", colour = "Temperature") +
    theme_classic2()
)
```

#3 Interpretation:

Distribution of plots across space, showing precipitation and temperature

higher precipitation in the west (lower longitude) plots, low in the north

#### Variations of trait against a climate space plots

- impact of temperature on traits

- impact of precipitation on traits

```{r plots_trait-temperature}

# Test plots of trait against a climate space

#1 plant_height/temperature + hovedtype
(plot_height_temp <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = plant_height_cm,
                   colour = hovedtype)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Plant Height (cm)", colour = "Vegetation Type") +
    theme_classic2()
)

#2 plant_height/temperature + species
(plot_height_temp2 <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = plant_height_cm,
                   colour = species)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Plant Height (cm)", colour = "Species") +
    theme_classic2()
)

#2b boxplot plant_height/temperature + species
(boxplot_height_temp2 <- ggplot(data = Gradient.df3) +
    geom_boxplot(aes(x = temperature_WQ,
                     y = plant_height_cm,
                     colour = species)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Plant Height (cm)", colour = "Species") +
    theme_classic2()
)

#3 length_lb/temperature + hovedtype
(plot_length_temp <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = length_lb_cm,
                   colour = hovedtype)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Longest Branch Length (cm)", colour = "Vegetation Type") +
    theme_classic2()
)

#4 length_lb/temperature + species
(plot_length_temp2 <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = length_lb_cm,
                   colour = species)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Longest Branch Length (cm)", colour = "Species") +
    theme_classic2()
)

# leaf_weight/temperature + hovedtype
(plot_leafweight_temp <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = leaf_weight,
                   colour = hovedtype)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Leaf Weight (g)", colour = "Vegetation Type") +
    theme_classic2()
)

# check to see if all samples that are weighed so far were included
#sum(is.na(Gradient.df3$leaf_weight)) # 996 NAs, explains most of the removed rows
#sum(is.na(Gradient.df3$hovedtype)) # 904
#sum(is.na(Gradient.df3$hovedtype) & is.na(Gradient.df3$leaf_weight)) # 875 overlap
# only data either missing hovedtype and/or not yet weighed was omitted, all good
# no need to run this part every time

# leaf_weight/temperature + species
(plot_leafweight_temp2 <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = leaf_weight,
                   colour = species)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Leaf Weight (g)", colour = "Species") +
    theme_classic2()
)

```

Initial observations on impact of temperature on traits: increase height and leaf weight

#2 and #2b VM plants tend to be a bit taller than VV

#3 and #4 show some VM plants from Fastmarksskogmark have especially long branches

\# one outlier in leaf weight!

```{r plots_trait-precipitation}

# same plots as above with precipation instead of temperature

#1 plant_height/precipitation + hovedtype
(plot_height_prec <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = precipitation_Annual,
                   y = plant_height_cm,
                   colour = hovedtype)) +
    labs(x = "Annual Precipitation", y = "Plant Height (cm)", colour = "Vegetation Type") +
    theme_classic2()
)

#2 plant_height/precipitation + species
(plot_height_prec2 <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = precipitation_Annual,
                   y = plant_height_cm,
                   colour = species)) +
    labs(x = "Annual Precipitation", y = "Plant Height (cm)", colour = "Species") +
    theme_classic2()
)

#3 length_lb/precipitation + hovedtype
(plot_length_prec <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = precipitation_Annual,
                   y = length_lb_cm,
                   colour = hovedtype)) +
    labs(x = "Annual Precipitation", y = "Longest Branch Length (cm)", colour = "Vegetation Type") +
    theme_classic2()
)

#4 length_lb/precipitation + species
(plot_length_prec2 <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = precipitation_Annual,
                   y = length_lb_cm,
                   colour = species)) +
    labs(x = "Annual Precipitation", y = "Longest Branch Length (cm)", colour = "Species") +
    theme_classic2()
)

# leaf_weight/precipitation + hovedtype
(plot_leafweight_prec <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = precipitation_Annual,
                   y = leaf_weight,
                   colour = hovedtype)) +
    labs(x = "Annual Precipitation", y = "Leaf Weight (g)", colour = "Vegetation Type") +
    theme_classic2()
)


# leaf_weight/precipitation + species
(plot_leafweight_prec2 <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = precipitation_Annual,
                   y = leaf_weight,
                   colour = species)) +
    labs(x = "Annual Precipitation", y = "Leaf Weight (g)", colour = "Species") +
    theme_classic2()
)


```

Patterns much less clear with precipitation than with temperature

Plant size traits seem to peek around 1200 mm annual precipitation, except for the especially tall VM plants from Fastmarksskogmark

#### Relationships between traits

```{r plots_trait-relationships}

```

## PCA

Are plants from the same hovedtype or the same location more similar to each other?

-\> within-group similarity

as an example I will use plant_height_cm, length_lb_cm and nr_segments_lb as the numerical variables

PCA to see if there is clustering by species, hovedtype or location

```{r PCA_attempt}

library(vegan)

PCA_data <- Gradient.df3 %>% select(plant_height_cm,length_lb_cm, nr_segments_lb) # select the numeric columns to use 
PCA_data_scaled <- decostand(PCA_data, method = "standardize")

PCA <- rda(PCA_data_scaled, scale = TRUE) # true because variables have different units

summary(PCA) # PC1 explains about 70% of the variance # PC2 about 26%

site_scores <- as.data.frame(scores(PCA, display = "sites")) 
species_scores <- as.data.frame(scores(PCA, display = "species"))

PCA_df <- site_scores %>% mutate( location = Gradient.df3$ano_punkt,
                                  hovedtype = Gradient.df3$hovedtype,
                                  species = Gradient.df3$species) 
# connected PC1 and PC2 with the categories we want to compare

#now we plot PC1 vs PC2 with our categories, e.g. species
ggplot(PCA_df, aes(x = PC1, y = PC2, colour = species))+
  geom_point()+ 
  theme_minimal()+
  theme(legend.position = "right")+
  stat_ellipse(level = 0.95)

```

Interpretation: clustering by species fairly clear for VV

Also tried the same code with hovedtype -\> clustering not very strong, but that may change if more or different numerical measurements were used as a basis for the PCA
