---
title: "DURIN_Gradient_Cleaning_Climate_Sara"
author: "Sonya R. Geange"
format: html
editor: visual
---

## Introduction to Project

Can add a few lines here about what this script is trying to achieve.

## Install and load packages, Import Data

```{r load_packages}
#| message: false

# Load the required libraries

# Toinstall packages the first time, run this code
# install.packages("name")

# once installed, run this code
library(tidyverse)
library(readxl)
library(ggplot2)
library(grid)
library(xlsx)
library(RColorBrewer)
library(lme4)
library(lmerTest)
library(emmeans)
library(tidyr)
library(ggthemes)
library(ggpubr)
```

Import data

```{r import_data}

# Import the dataset for the field and lab
Gradient.df <- read_excel("C:/Users/sarag/Uni/DURIN Gradient Project/data/raw_data/Gradient sampling 2024 - DurinV5.xlsx",
                          col_types = c("guess", "text", "text", "text",
                                        "text", "text", "text", "numeric",
                                        "numeric", "numeric", "numeric", "text",
                                        "numeric", "numeric", "numeric", "numeric",
                                        "numeric","numeric", "numeric", "numeric",
                                        "numeric", "numeric", "numeric", "text",
                                        "text", "text", "text", "text", "text", "text", "text", "text"))
Gradient.dfv2 <- read_csv("C:/Users/sarag/Uni/DURIN Gradient Project/data/raw_data/Gradient sampling 2024 - DurinV5.csv", locale=locale(encoding="latin1")
                          )
# the latin1 encoding introduced a problem here: added names to empty columns, need to be removed
Gradient.dfv2 <- Gradient.dfv2 %>% 
  select(-...33, -...34, -...35, -...36, -...37)

# Import ANO site and climate data
all_coords <- read.csv("C:/Users/sarag/Uni/DURIN Gradient Project/data/raw_data/all_coords.csv")

# Sampled ANO sites
ANO_sites <- read_excel("C:/Users/sarag/Uni/DURIN Gradient Project/data/raw_data/Gradient sampling 2024 - DurinV5.xlsx",
                        sheet = "ANOSampledSites",
                          col_types = c("guess", "text", "text", "text", "text"))

```

## Data cleaning

```{r four_corners}

gradient.four.corner.df <- Gradient.dfv2 %>%
  # Drop fill the other columns to complete meta-data
  fill(date, .direction = "down") %>%
  fill(ANO_flatID, .direction = "down") %>%
  # Filter to the DURIN 4 Corner sites
  filter(ANO_flatID %in% c("SE_O",
                           "SE_F",
                           "KA_O",
                           "KA_F",
                           "SO_F",
                           "SO_O",
                           "LY_O",
                           "LY_F",
                           "KK_O",
                           "KK_F")) %>%
  # Drop fill only the key columns to help split the dataset
  fill(ANO_pointID, .direction = "down") %>%
  fill(habitat, .direction = "down") %>%
  fill(collector, .direction = "down") %>%
  fill(NiN_type, .direction = "down") %>%
  fill(species, .direction = "down") %>%
  fill(individual_nr, .direction = "down") %>%
  fill(plant_height_cm, .direction = "down") %>%
  fill(length_lb_cm, .direction = "down") %>%
  fill(nr_segments_lb, .direction = "down")  %>%
  # Making sure that the data columns are numeric, Segment column as character
  mutate(across(c(plant_height_cm,
                  length_lb_cm,
                  nr_segments_lb),
                as.numeric)) %>%
  mutate(across(c(
    length_mm,
    nr_leaves,
    nr_scars,
    nr_branch_scars,
    nr_side_branches,
    nr_fruits,
    leaf_weight,
    stem_weight,
    fruit_weight,
    branches_weight,
    sidebranches_weight),
    as.numeric)) %>%
  mutate(across(c(Segment),
                as.character)) %>%
  # Change the spelling of 'KK_F' and 'KK_O' to match correct format
  # During processing KA's and KK's were accidnetly duplicated during data entry of field information
  # Only KK's should be kept as these have field and lab data (weights)
  filter(ANO_flatID %in% c("KK_O", "KK_F",
                           "SE_O", "SE_F",
                           "SO_O", "SO_F",
                           "LY_O", "LY_F")) %>%
  # Rename KK's to KA's to keep naming structure
  mutate(ANO_flatID = recode(ANO_flatID,
                             'KK_F'='KA_F',
                             'KK_O' = 'KA_O')) %>%
  # Change spelling of 'open' to 'Open' to match DURIN formatting standards
  mutate(habitat = recode(habitat,
                          'open'='Open',
                          'forest'='Forest')) %>%
  # Create a new siteID column, and use info in ANO_flatID to allocate sites
  mutate(siteID = recode(ANO_flatID,
                         'SE_O' = 'Senja',
                         'SE_F' = 'Senja',
                         'KA_O' = 'Kautokeino',
                         'KA_F' = 'Kautokeino',
                         'SO_F' = 'Sogndal',
                         'SO_O' = 'Sogndal',
                         'LY_O' = 'Lygra',
                         'LY_F' = 'Lygra')) %>%
  # Change spelling of species and column name to match DURIN data standards
  mutate(speciesID = recode(species,
                            'vv' = 'VV',
                            'vm' = 'VM')) %>%
  # Create column for full species names, and match DURIN data standards
  mutate(species = recode(species,
                          'vv'= 'Vaccinium vitis-idaea',
                          'vm' = 'Vaccinium myrtillus')) %>%
  # Add context for the sites (e.g. Lygra is in the South and by the Coast)
  mutate(geography = factor(siteID, levels = c("Lygra", "Senja", "Sogndal", "Kautokeino"),
                            labels = c("South", "North", "South", "North"))) %>%
  mutate(oceanity = factor(siteID, levels = c("Lygra", "Senja", "Sogndal", "Kautokeino"),
                           labels = c("Coast", "Coast", "Inland", "Inland"))) %>% 
  mutate(
    leaf_stem_ratio_main = leaf_weight / stem_weight,
    total_biomass = rowSums(select(., leaf_weight:sidebranches_weight), na.rm = TRUE))
# add a new column for Leaf/stem ratio and total biomass to be calculated based on existing weight column content
```

```{r climate}

# Create the climate dataset
# Tidy up the dataset names and values
Gradient.df2 <- Gradient.dfv2 %>%
  # Drop fill only the key columns to help split the dataset
  # Date, the ANO_flatID which is a mix of climate, 4corners and VCG coding and in some instance ANO + pointID
  fill(date, .direction = "down") %>%
  fill(ANO_flatID, .direction = "down") %>%
  # filter out (remove) four corner sites
  filter(! ANO_flatID %in% c("SE_O", "SE_F",
                             "KA_O", "KA_F",
                             "KK_O", "KK_F",
                             "LY_O", "LY_F",
                             "SO_O", "SO_F")) %>%
  # Also drop Mika's sites from VCG
  filter(! ANO_pointID %in% c("MK8_4", "MK9_2", "MK8_2",
                              "MK19_1", "MK8_3", "MK8_1",
                              "MK9_4", "MK9_3")) %>%
  # And the sites from Finland (no coords) %>%
  filter(! ANO_flatID %in% c("KIL")) %>%
  # String split ANO_flatID as some instances have the pointID also included
  separate_wider_delim(cols = ANO_flatID,
                       delim = '_',
                       names = c('ANO_flatID', 'ANO_pointID2'),
                       too_few = "align_start") %>%
  # Merge the original ANO_pointID with the newly populated ANO_point2
  mutate(ANO_pointID = coalesce(ANO_pointID, ANO_pointID2)) %>%
  # Drop fill the rest of the columns now the ANO ID's are correct
  fill(ANO_pointID, .direction = "down") %>%
  fill(habitat, .direction = "down") %>%
  fill(collector, .direction = "down") %>%
  fill(NiN_type, .direction = "down") %>%
  fill(species, .direction = "down") %>%
  fill(individual_nr, .direction = "down") %>%
  fill(plant_height_cm, .direction = "down") %>%
  fill(length_lb_cm, .direction = "down") %>%
  fill(nr_segments_lb, .direction = "down")  %>%
  # For external collaborators
  fill(UniqueID, .direction = "down")  %>%
  # Make these columns numeric
  mutate(across(c(
    # Change the plotID codes back to numeric for later on
    ANO_flatID,
    ANO_pointID,
    # Other normal variables
    plant_height_cm,
    length_lb_cm,
    nr_segments_lb,
    length_mm,
    nr_leaves,
    nr_scars,
    nr_branch_scars,
    nr_side_branches,
    nr_fruits,
    leaf_weight,
    stem_weight,
    fruit_weight,
    branches_weight,
    sidebranches_weight),
    as.numeric)) %>%
  # Change spelling of 'open' to 'Open' to match DURIN formatting standards
  mutate(habitat = recode(habitat,
                          'skog' = 'Forest',
                          'open'='Open',
                          'forest'='Forest')) %>%
  # For consistency with 4corners
  # Create a new siteID column, and use info in ANO_flatID to allocate sites
  add_column(siteID = NA) %>%
  # Change spelling of species and column name to match DURIN data standards
  mutate(speciesID = recode(species,
                            'vv' = 'VV',
                            'vm' = 'VM')) %>%
  # Create column for full species names, and match DURIN data standards
  mutate(species = recode(species,
                          'vv'= 'Vaccinium vitis-idaea',
                          'vm' = 'Vaccinium myrtillus')) %>%
  # Remove columns not needed
  select(!ANO_pointID2) %>%
  # Test create new ANO_flatID column removing additional 0 in front
  mutate(ANO_flatID2 = as.numeric(str_remove(ANO_flatID, "^0+")))
```

```{r climate_ANO_IDs}

Gradient.df2 <- Gradient.df2 %>%
  # Next issue is variable ways that the ANO_flatID has been entered (number of digits)
  # With 2, 3 and 4 letter codes we change the Prefix we add
  # With the values above 1000, we don't want the 0's at all
  mutate(ano_punkt = case_when(
    ANO_flatID2 > 0 & ANO_flatID2 <= 99 ~ paste0("ANO00", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    ANO_flatID2 >= 100 & ANO_flatID2 <= 999 ~ paste0("ANO0", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    ANO_flatID2 >= 1000 ~ paste0("ANO", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    TRUE ~ NA))
```

The following block is unfinished, does not run

```{r sites_ANO_IDs}
#| eval: false

# ANO Sites
ANO_sites2 <-  ANO_sites %>%  # Test create new ANO_flatID column removing additional 0 in front
  mutate(ANO_flatID2 = as.numeric(str_remove(ANO_flatID, "^0+")))
# Next issue is variable ways that the ANO_flatID has been entered (number of digits)
# With 2, 3 and 4 letter codes we change the Prefix we add
# With the values above 1000, we don't want the 0's at all
ANO_sites2 <- ANO_sites2 %>% 
  mutate(ano_punkt = case_when(
    ANO_flatID2 > 0 & ANO_flatID2 <= 99 ~ paste0("ANO00", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    ANO_flatID2 >= 100 & ANO_flatID2 <= 999 ~ paste0("ANO0", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    ANO_flatID2 >= 1000 ~ paste0("ANO", Gradient.df2$ANO_flatID2, "_", Gradient.df2$ANO_pointID),
    TRUE ~ NA))
##########################
#TODO:
# Edit Pål's rows, as all plants are listed as individual 1 atm, whereas we need to use the unique ID column to change this.
# Kristine sent through the dataset with the uniqueID, so cross-reference against that

# For LGV and PT, need to check branch numbers vs branch scars.
# For AH, KB, SG, and maybe MK we counted the number of branches removed (base of branch) from the main stem.
# However, LGV and PT counted the number of tips on the branches removed.

# LGV and PT leaf scars vs leaf counts. They have leaf scars in addition to the leaf counts
# AH, KB, SG and maybe MK have removed leaves and counted, then counted all scars remaining on the branch. 
```

```{r climate_ANO}

# Update ano_punkt so only climate gradient sites exist (as they can be matched)
# So all sites with forest/open design (i.e. 4corners and VCG are as NA's, and Finland samples)
Gradient.df2 <- Gradient.df2 %>%
  mutate(ano_punkt = case_when(
    habitat %in% c("Open", "Forest") ~ NA,
    .default = as.character(ano_punkt)
  ))

# Filter the ANO plot co-ordinate and climate data to keep only distinct records for each location
all_coords_distinct <- all_coords %>%
  distinct(ano_punkt, .keep_all = TRUE)

# Create a dataframe just for the sites with coordinates for climate plotting
Gradient.df3 <- Gradient.df2 %>%
  left_join(all_coords_distinct, join_by(ano_punkt)) %>% 
  mutate(
    leaf_stem_ratio_main = leaf_weight / stem_weight,
    total_biomass = rowSums(select(., leaf_weight:sidebranches_weight), na.rm = TRUE))


# Create a dataframe for just the ANO sampled sites, and link to ANO climate data
ANO_site_climate <- ANO_sites %>%
  left_join(all_coords_distinct, join_by(ano_punkt))

```

Names of vegetation types

```{r cleaning_vegetation_types}

#unique(Gradient.df3$hovedtype)
#class(Gradient.df3$hovedtype)
Gradient.df3$hovedtype <- as.factor(Gradient.df3$hovedtype)
levels(Gradient.df3$hovedtype)

Gradient.df3$hovedtype <- factor(Gradient.df3$hovedtype, 
                                 labels = c('Åpen grunnlendt mark', 'Åpen jordvannsmyr', 'Boreal hei', 'Fastmarksskogsmark', 'Fjellhei leside og tundra', 'Grus- og steindominert strand og strandlinje', 'Myr- og sumpskogsmark', 'Nakent berg', 'Nedbørsmyr', 'Rabbe', 'Semi-naturlig eng', 'Skogsmark'))
# renaming 
# having set file encoding option to "latin1" in the beginning when first importing the data helps R interspret the special letters correctly

Gradient.df3$hovedtype <- factor(Gradient.df3$hovedtype,
                                 levels = c('Nakent berg', 'Åpen grunnlendt mark', 'Fjellhei leside og tundra', 'Fastmarksskogsmark', 'Skogsmark', 'Rabbe', 'Grus- og steindominert strand og strandlinje', 'Boreal hei', 'Semi-naturlig eng', 'Åpen jordvannsmyr', 'Myr- og sumpskogsmark', 'Nedbørsmyr'))
# reordering by NiN types, T for Fastmarkssystemer and V for Våtmarkssystemer 

## should Skogsmark also be called Fastmarksskogsmark?
Gradient.df3 <- Gradient.df3 %>% 
  mutate(hovedtype = recode(hovedtype, 'Skogsmark'='Fastmarksskogsmark'))
```

## Plots

```{r plots_setup}

## Define colour palette (used with scale fill manual)
Habitat <- c("Forest" = "#083508", "Open" = "#589758")
Species <- c("Vaccinium myrtillus" = "#087e8b",   # berry colour option "#323284"
             "Vaccinium vitis-idaea" =  "#eb5461") # berry colour option "#D93137"
siteID <- c("Lygra" = "#4A235A",
            "Sogndal" = "#154360",
            "Senja" = "#C39BD3",
            "Kautokeino" = "#5499C7")
```

#### Plots on sites and climate space

1.  Vegetation types across temperature and precipitation gradient

2.  Vegetation types across temperature and precipitation gradient, aesthetic changes

3.  Temperature and precipitation across space (coordinates)

```{r test_plots}

#1 Do a test plot of climate space we sampled across
(plot <- ggplot(data = Gradient.df3) +
    geom_jitter(aes(x = precipitation_Annual,
                    y = temperature_WQ,
                    colour = hovedtype),
                # Slightly offset the plots so its easier to visualize the samples per site
                height = 0.05, # change is in y-axis values (i.e. temperature)
                width = 25, # change is in x-axis values (i.e. precipitation)
                size = 2))


#2 Do a test plot of climate space we sampled across
(plot_ANO_climate <- ggplot(data = ANO_site_climate) +
    geom_jitter(aes(x = precipitation_Annual,
                    y = temperature_WQ,
                    colour = hovedtype),
                # Slightly offset the plots so its easier to visualize the samples per site
                height = 0.1, # change is in y-axis values (i.e. temperature)
                width = 25, # change is in x-axis values (i.e. precipitation)
                size = 2) +
    theme_classic() +
    labs(x = "Precipitation (annual, mm/yr)", y = "Temperature (Warmest Quarter)", colour = "Vegetation Type") )


#3 temperature and precipiation across space
(plot_space_climate <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = longitude,
                   y = latitude,
                   colour = temperature_WQ,
                   size = precipitation_Annual)) +
    scale_colour_gradient(low = "blue", high = "red")+
    labs(x = "Longitude", y = "Latitude", colour = "Temperature") +
    theme_classic2()
)
```

#3 Interpretation:

Distribution of plots across space, showing precipitation and temperature

higher precipitation in the west (lower longitude) plots, low in the north

#### Variations of trait against a climate space plots

##### Impact of temperature on traits

```{r plots_trait-temperature}

# Test plots of trait against a climate space

#1 plant_height/temperature + hovedtype
(plot_height_temp <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = plant_height_cm,
                   colour = hovedtype)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Plant Height (cm)", colour = "Vegetation Type") +
    theme_classic2()
)

#2 plant_height/temperature + species
(plot_height_temp2 <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = plant_height_cm,
                   colour = species)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Plant Height (cm)", colour = "Species") +
    theme_classic2()
)

#2b boxplot plant_height/species
(boxplot_height_temp2 <- ggplot(data = Gradient.df3) +
    geom_boxplot(aes(x = species,
                     y = plant_height_cm)) +
    labs(x = "Species", y = "Plant Height (cm)") +
    theme_classic2()
)

#3 length_lb/temperature + hovedtype
(plot_length_temp <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = length_lb_cm,
                   colour = hovedtype)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Longest Branch Length (cm)", colour = "Vegetation Type") +
    theme_classic2()
)

#4 length_lb/temperature + species
(plot_length_temp2 <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = length_lb_cm,
                   colour = species)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Longest Branch Length (cm)", colour = "Species") +
    theme_classic2()
)

#5 total_biomass/temperature + species
(plot_biomass_temp <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = total_biomass,
                   colour = species)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Total Biomass (2022-2024)", colour = "Species") +
    ylim(0.0000, 1.0000) +
    theme_classic2()
)
# VM seems to overall be heavier, but is this the case for each year's growth equally? 

# broken down by year
#5a 2024
Gradient.df3.2024 <- filter(Gradient.df3, Segment %in% c("S24")) #select only 2024

(plot_biomass_temp_2024 <- ggplot(data = Gradient.df3.2024) +
    geom_point(aes(x = temperature_WQ,
                   y = total_biomass,
                   colour = species)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Total Biomass 2024", colour = "Species") +
    ylim(0.0000, 1.0000) +
    theme_classic2()
)
# outlier, almost definitely a mistake (biomass 286g), exclude by redifining axis limits to make the plot useable

#5b 2023
Gradient.df3.2023 <- filter(Gradient.df3, Segment %in% c("S23")) #select only 2023

(plot_biomass_temp_2023 <- ggplot(data = Gradient.df3.2023) +
    geom_point(aes(x = temperature_WQ,
                   y = total_biomass,
                   colour = species)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Total Biomass 2023", colour = "Species") +
    theme_classic2()
)

#5c 2022
Gradient.df3.2022 <- filter(Gradient.df3, Segment %in% c("S22")) #select only 2022

(plot_biomass_temp_2022 <- ggplot(data = Gradient.df3.2022) +
    geom_point(aes(x = temperature_WQ,
                   y = total_biomass,
                   colour = species)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Total Biomass 2022", colour = "Species") +
    theme_classic2()
)


# leaf_weight/temperature + hovedtype
(plot_leafweight_temp <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = leaf_weight,
                   colour = hovedtype)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Leaf Weight (g)", colour = "Vegetation Type") +
    theme_classic2()
)

# check to see if all samples that are weighed so far were included
#sum(is.na(Gradient.df3$leaf_weight)) # 996 NAs, explains most of the removed rows
#sum(is.na(Gradient.df3$hovedtype)) # 904
#sum(is.na(Gradient.df3$hovedtype) & is.na(Gradient.df3$leaf_weight)) # 875 overlap
# only data either missing hovedtype and/or not yet weighed was omitted, all good
# no need to run this part every time

# leaf_weight/temperature + species
(plot_leafweight_temp2 <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = leaf_weight,
                   colour = species)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Leaf Weight (g)", colour = "Species") +
    theme_classic2()
)

```

Initial observations on impact of temperature on traits: size and weight tend to increase with temperature

#2 and #2b show that VM plants tend to be a bit taller than VV, possible outliers in VM

#3 and #4 show some VM plants from Fastmarksskogmark have especially long branches

\# one outlier in weight: 1427_46_VV_2 in 2024 leaf weight (and total biomass as a consequence) -\> 286g seems unrealistic, likely a typo

#5c VM plants were heavier than VV in 2022, much more noticable than any patterns in the following years -\> species differences more closely analysed in the next code block

##### Impact of precipitation on traits

```{r plots_trait-precipitation}

# same plots as above with precipation instead of temperature

#1 plant_height/precipitation + hovedtype
(plot_height_prec <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = precipitation_Annual,
                   y = plant_height_cm,
                   colour = hovedtype)) +
    labs(x = "Annual Precipitation", y = "Plant Height (cm)", colour = "Vegetation Type") +
    theme_classic2()
)

#2 plant_height/precipitation + species
(plot_height_prec2 <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = precipitation_Annual,
                   y = plant_height_cm,
                   colour = species)) +
    labs(x = "Annual Precipitation", y = "Plant Height (cm)", colour = "Species") +
    theme_classic2()
)

#3 length_lb/precipitation + hovedtype
(plot_length_prec <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = precipitation_Annual,
                   y = length_lb_cm,
                   colour = hovedtype)) +
    labs(x = "Annual Precipitation", y = "Longest Branch Length (cm)", colour = "Vegetation Type") +
    theme_classic2()
)

#4 length_lb/precipitation + species
(plot_length_prec2 <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = precipitation_Annual,
                   y = length_lb_cm,
                   colour = species)) +
    labs(x = "Annual Precipitation", y = "Longest Branch Length (cm)", colour = "Species") +
    theme_classic2()
)

# leaf_weight/precipitation + hovedtype
(plot_leafweight_prec <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = precipitation_Annual,
                   y = leaf_weight,
                   colour = hovedtype)) +
    labs(x = "Annual Precipitation", y = "Leaf Weight (g)", colour = "Vegetation Type") +
    theme_classic2()
)


# leaf_weight/precipitation + species
(plot_leafweight_prec2 <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = precipitation_Annual,
                   y = leaf_weight,
                   colour = species)) +
    labs(x = "Annual Precipitation", y = "Leaf Weight (g)", colour = "Species") +
    theme_classic2()
)


```

Patterns much less clear with precipitation than with temperature

Plant size traits seem to peek around 1200 mm annual precipitation, except for the especially tall VM plants from Fastmarksskogmark

#### Function to automatically create multiple variations of these plots

Automatic plots for traits against temperature, colour showing hovedtype

```{r auto_plot_trait/temperature1}

# Base-plot: plant_height/temperature + hovedtype
(plot_height_temp <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = plant_height_cm,
                   colour = hovedtype)) +
    labs(x = "Temperature (Warmest Quarter)", y = "Plant Height (cm)", colour = "Vegetation Type") +
    theme_classic2()
)

# Function setup
trait_temperature_plots <- function(trait_value){
  plot_trait_value_temp <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = .data[[trait_value]],  #dynamic column selection
                   colour = hovedtype)) +
    labs(x = "Temperature (Warmest Quarter)", colour = "Vegetation Type") +
    theme_classic2()
}

(trait_temperature_plots("plant_height_cm")) #Test: now it works!

trait_temperature_plotlist = list()

for (i in c("plant_height_cm", "length_lb_cm", "length_mm", "nr_leaves", "nr_branch_scars", "nr_side_branches", "leaf_weight", "stem_weight", "branches_weight", "leaf_stem_ratio_main", "total_biomass" )) {trait_temperature_plotlist[[i]] = trait_temperature_plots(i)}

trait_temperature_plotlist
```

Automatic plots for traits against temperature, colour showing species

```{r auto_plot_trait/temperature2}

# Function setup
trait_temperature_plots2 <- function(trait_value){
  plot_trait_value_temp2 <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = .data[[trait_value]],  #dynamic column selection
                   colour = species)) +
    labs(x = "Temperature (Warmest Quarter)", colour = "Species") +
    theme_classic2()
}

trait_temperature_plotlist2 = list()

for (i in c("plant_height_cm", "length_lb_cm", "length_mm", "nr_leaves", "nr_branch_scars", "nr_side_branches", "leaf_weight", "stem_weight", "branches_weight", "leaf_stem_ratio_main", "total_biomass" )) {trait_temperature_plotlist2[[i]] = trait_temperature_plots2(i)}

trait_temperature_plotlist2


```

Automatic plots for traits against precipitation, colour showing hovedtype

```{r auto_plot_trait/precipitation1}

# Function setup
trait_precipitation_plots <- function(trait_value){
  plot_trait_value_prec <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = precipitation_Annual,
                   y = .data[[trait_value]],  #dynamic column selection
                   colour = hovedtype)) +
    labs(x = "Annual Precipitation", colour = "Vegetation Type") +
    theme_classic2()
}

trait_precipitation_plotlist = list()

for (i in c("plant_height_cm", "length_lb_cm", "length_mm", "nr_leaves", "nr_branch_scars", "nr_side_branches", "leaf_weight", "stem_weight", "branches_weight", "leaf_stem_ratio_main", "total_biomass" )) {trait_precipitation_plotlist[[i]] = trait_precipitation_plots(i)}

trait_precipitation_plotlist
```

Automatic plots for traits against precipitation, colour showing species

```{r auto_plot_trait/precipitation2}

# Function setup
trait_precipitation_plots2 <- function(trait_value){
  plot_trait_value_prec2 <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = precipitation_Annual,
                   y = .data[[trait_value]],  #dynamic column selection
                   colour = species)) +
    labs(x = "Annual Precipitation", colour = "Species") +
    theme_classic2()
}

trait_precipitation_plotlist2 = list()

for (i in c("plant_height_cm", "length_lb_cm", "length_mm", "nr_leaves", "nr_branch_scars", "nr_side_branches", "leaf_weight", "stem_weight", "branches_weight", "leaf_stem_ratio_main", "total_biomass" )) {trait_precipitation_plotlist2[[i]] = trait_precipitation_plots2(i)}

trait_precipitation_plotlist2
```

For any traits that don't apply to the whole plant it would make sense to look at each year individually

```{r by_segment}

(plot_leafweight2024 <- Gradient.df3 %>% 
    filter(Segment == "S24", 
          leaf_weight <0.5) %>%  #removes outlier
    ggplot(data = .) +
    geom_point(aes(x = temperature_WQ,
                   y = leaf_weight,
                   colour = species)) +
    theme_classic2()
)


#comparing between years
(boxplot_stemweight <- Gradient.df3 %>% group_by(Segment) %>% 
    filter(!is.na(Segment), stem_weight <10) %>% 
    ggplot(data = .) +
    geom_boxplot(aes(x = Segment,
                     y = stem_weight,
                     fill = Segment)) +
    theme_classic2()
)
# not the most helpful looking boxplots
```

Looking for Outliers and Issues in the data

```{r outliers}

# stem_weight for 1427_46_vv_2 = 286
# this affects stem_weight, leaf_stem_ratio_main, total_biomass

# leaf weight shows one single value above 0.5, might be an outlier
Gradient.df3[Gradient.df3$leaf_weight > 0.5, ] 
# leaf_weight = 0.9709 for 622_35_T4-C-1_vm_1 (5 leaves)
# is this still a realisitic value for leaf weight?

# number of side branches potocol issue with counting tips vs from the base of each branch

```

Code to modify individual plots to exclude these outliers or make other modifications:

The outlier in stem weight for example affects all plots showing stem_weight, total_biomass and leaf_stem_ratio_main (though in a more subtle way, doesn't visually stand out here)

The possible outlier in leaf weight impacts plots showing leaf_weight, total_biomass and leaf_stem_ratio_main

```{r modify_plots}

# exclude by adjusting axis limits (won't work for leaf_stem_ratio)
for (i in c("stem_weight", "total_biomass")){
  trait_temperature_plotlist[[i]] <- trait_temperature_plotlist[[i]]+
    ylim(0, 2)
}

trait_temperature_plotlist[["stem_weight"]]
trait_temperature_plotlist[["total_biomass"]]

# same for leaf_weight outlier
for (i in c("leaf_weight", "total_biomass")){
  trait_temperature_plotlist[[i]] <- trait_temperature_plotlist[[i]]+
    ylim(0, 0.5)
}

trait_temperature_plotlist[["leaf_weight"]]
trait_temperature_plotlist[["total_biomass"]]

# to handle these outliers in other contexts too, it might be better to filter them out of the dataframe, instead of relying on the axes

```

#### Plots on species differences

```{r plots_species differences}

# VM seems to generally be heavier, is this true for all seasons/traits?

# VM plants were heavier than VV in 2022 (noticed from plot #5c in previous chunk), much more noticable than any patterns in the following years 

# is this due to the segments being longer (stem_weight) or the number and weight of side branches (nr_side_branches and branches_weight)?

#1a stem_weight(2022)/species boxplot
(boxplot_stemweight_species_2022 <- ggplot(data = Gradient.df3.2022) +
    geom_boxplot(aes(x = species,
                     y = stem_weight, colour = species)) +
    labs(x = "Species", y = "Stem weight (g) 2022") +
    theme_classic2()
)

#1b branches_weight(2022)/species
(boxplot_branchweight_species_2022 <- ggplot(data = Gradient.df3.2022) +
    geom_boxplot(aes(x = species,
                     y = branches_weight, colour = species)) +
    labs(x = "Species", y = "Weight of Side Branches (g) 2022") +
    theme_classic2()
)
# the difference between the species looks more pronounced for the weight of side branches, though the pattern of VM being heavier persists in stem weight too
# is that because VM had more side branches in 2022?

#1c nr_side_branches/species
(boxplot_branchnr_species_2022 <- ggplot(data = Gradient.df3.2022) +
    geom_boxplot(aes(x = species,
                     y = nr_side_branches, colour = species)) +  
    labs(x = "Species", y = "Number of Side Branches 2022") +
    theme_classic2()
)
# issues with the protocol for counting branches (counting tips)
unique(Gradient.df3.2022$nr_side_branches)
unique(Gradient.df3.2022$nr_branch_scars) # the max number of side branches should not exceed the number of branch scars (13)
# due to frequent incorrect counts, using branch scar count as a proxy for number of side branches for now (note that sometimes not all side branches were actually present, so corrected counts are preferable)
#1d branch_scars/species
(boxplot_branchscar_species_2022 <- ggplot(data = Gradient.df3.2022) +
    geom_boxplot(aes(x = species,
                     y = nr_branch_scars, colour = species)) +  
    labs(x = "Species", y = "Number of Branch Scars 2022") +
    theme_classic2()
)
# much clearer picture, VM tends to branch more
# need to consider this when comparing biomass 

# does this pattern persist in 2023?

#2a stem_weight(2023)/species boxplot
(boxplot_stemweight_species_2023 <- ggplot(data = Gradient.df3.2023) +
    geom_boxplot(aes(x = species,
                     y = stem_weight, colour = species)) +
    labs(x = "Species", y = "Stem weight (g) 2023") +
    theme_classic2()
)

#2b branches_weight(2022)/species
(boxplot_branchweight_species_2023 <- ggplot(data = Gradient.df3.2023) +
    geom_boxplot(aes(x = species,
                     y = branches_weight, colour = species)) +
    labs(x = "Species", y = "Weight of Side Branches (g) 2023") +
    theme_classic2()
)

# skipping 2c (nr_side_branches/species) for now

#2d branch_scars/species
(boxplot_branchscar_species_2023 <- ggplot(data = Gradient.df3.2023) +
    geom_boxplot(aes(x = species,
                     y = nr_branch_scars, colour = species)) +  
    labs(x = "Species", y = "Number of Branch Scars 2023") +
    theme_classic2()
)
# pattern persists or even looks stronger in 2023, this is a general species difference, not something unique to the 2022 season

#3 boxplots for total biomass by year and species
(boxplot_biomass_year <- ggplot(data = Gradient.df3) +
    geom_boxplot(aes(x = Segment,
                     y = total_biomass, 
                     colour = species)) +
    labs(x = "Season", y = "Total Biomass (g)") +
    ylim(0.0000, 1.0000) + # exclude the 286g outlier from 2024 (VV)
    theme_classic2()
  )

# now compare leaf weight between species for 2024 (VM is deciduous, so no leaves from previous seasons)
#4a boxplot for leaf weight by species
(boxplot_leaf_weight_2024 <- ggplot(data = Gradient.df3.2024) +
    geom_boxplot(aes(x = species,
                     y = total_biomass, 
                     colour = species)) +
    labs(x = "Species", y = "Total Biomass (g)") +
    theme_classic2() + 
    ylim(0.0000, 1.0000)
)

#4b boxplot for leaf number (2024) by species 
(boxplot_leaf_number_2024 <- ggplot(data = Gradient.df3.2024) +
    geom_boxplot(aes(x = species,
                     y = nr_leaves, 
                     colour = species)) +
    labs(x = "Species", y = "Leaf Number") +
    theme_classic2() 
)
# VV tends to have more leaves 
```

#1c Protocol issue with counting side branches (tip counting in some, base counting in others) needs to be resolved to make all datapoints comparable

#2d and #2d VM branches more, need to be aware of this difference when comparing between species

#3 and previous: Weight differences between VM and VV persist for main stem and side branches, VM is consistently heavier

#4 higher leaf number in VV, overall leaf weight similar on average

#### Relationships between traits

```{r plots_trait-relationships}

# is number of segments (proxy for age) related to plant height?
(plot_height_segments <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = nr_segments_lb,
                   y = plant_height_cm,
                   colour = species)) +
    labs(x = "Number of Segments", y = "Plant Height (cm)", colour = "Species") +
    theme_classic2()
)
# slight trend of older plants being taller, but there are many older plants with very short segments
# more of an upper end (plants can only get so tall with a certain number of segments), but many short segments are common

# same with longest branch length instead of plant height
(plot_length_segments <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = nr_segments_lb,
                   y = length_lb_cm,
                   colour = species)) +
    labs(x = "Number of Segments", y = "Longest Branch (cm)", colour = "Species") +
    theme_classic2()
)
# shows a very similar pattern
```

## Regression Analysis

Testing initial observation: Many trait values (size and weight) increase with temperature.

Attempt to use a linear model (ran out of time to work out all the issues/try more variations, maybe a glm would be better)

```{r regression}

# temperature as a predictor of plant height
model_temp_height <- lm(plant_height_cm ~ temperature_WQ, data = Gradient.df3)
summary(model_temp_height)
# significant, explains about 25% of the variation in plant height

#visualisation
(plot_model_temp_height <- ggplot(data = Gradient.df3) +
    geom_point(aes(x = temperature_WQ,
                   y = plant_height_cm,
                   colour = hovedtype)) +
    stat_smooth(method = "lm", aes(temperature_WQ, plant_height_cm)) + 
    labs(x = "Temperature (Warmest Quarter)", y = "Plant Height (cm)", colour = "Vegetation Type") +
    theme_classic2()
)

plot(model_temp_height) 

model1_resids <- resid(model_temp_height)
shapiro.test(model1_resids)  # significant, non-normal distribution of residuals, likely due to some outliers

library(car)
model1_bptest <- ncvTest(model_temp_height)
print(model1_bptest)  # significant, have to assume heteroscedacity!



# lets add segment count as a proxy for age into this  model
Gradient.df3$nr_segments_lb <- as.numeric(Gradient.df3$nr_segments_lb)
model_temp_height2 <- lm(plant_height_cm ~ temperature_WQ + nr_segments_lb, data = Gradient.df3)
summary(model_temp_height2)

# better, both of these variables explain almost 30% of variation in plant height
```

Are there any such patterns for traits and precipitation?

## PCA

Are plants from the same hovedtype or the same location more similar to each other?

-\> within-group similarity

as an example I will use plant_height_cm, length_lb_cm and nr_segments_lb as the numerical variables

PCA to see if there is clustering by species, hovedtype or location

```{r PCA_attempt}

library(vegan)

PCA_data <- Gradient.df3 %>% select(plant_height_cm,length_lb_cm, nr_segments_lb) # select the numeric columns to use 
PCA_data_scaled <- decostand(PCA_data, method = "standardize")

PCA <- rda(PCA_data_scaled, scale = TRUE) # true because variables have different units

summary(PCA) # PC1 explains about 70% of the variance # PC2 about 26%

site_scores <- as.data.frame(scores(PCA, display = "sites")) 
species_scores <- as.data.frame(scores(PCA, display = "species"))

PCA_df <- site_scores %>% mutate( location = Gradient.df3$ano_punkt,
                                  hovedtype = Gradient.df3$hovedtype,
                                  species = Gradient.df3$species) 
# connected PC1 and PC2 with the categories we want to compare

#now we plot PC1 vs PC2 with our categories, e.g. species
ggplot(PCA_df, aes(x = PC1, y = PC2, colour = species))+
  geom_point()+ 
  theme_minimal()+
  theme(legend.position = "right")+
  stat_ellipse(level = 0.95)

```

Interpretation: clustering by species fairly clear for VV

Also tried the same code with hovedtype -\> clustering not very strong, but that may change if more or different numerical measurements were used as a basis for the PCA
